---
title: "R Exercises - Religion and babies"
author: "Sarvenaz Khodayari"
date: "10/10/2020"
output: html_document
---
`
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, echo= FALSE}
 library(tidyverse)
#children
children<- data.frame(read.csv("children.csv"))
#incomes
income<- data.frame(read.csv("income.csv"))
#population
pop<- data.frame(read.csv("pop.csv"))
#religion
religion<- data.frame(read.csv("religion.csv"),header= TRUE)
DF<- religion[,7:14]
religion$Religion<-max.religion<-colnames(DF)[max.col(DF)]


names(children)<- c("Country",c(1800:2018))
names(income)<- c("Country",c(1800:2040))
names(pop)<- c("Country",c(1800:2100))

```
```{r,echo= FALSE}
###PART1
library(tidyverse)
###Pivot_longer
dc<- children%>%
  select(Country,'1800':'2018')%>%
  pivot_longer(col=-Country,
               names_to="Year",
               values_to="Fertility",
               names_transform=list(Year = as.numeric))

               
di<-income%>%
  select(Country,'1800':'2018')%>%
  pivot_longer(col=-Country,
               names_to="Year",
               values_to="Income",
               names_transform=list(Year = as.numeric))

dp<-pop%>%
  select(Country,'1800':'2018')%>%
  pivot_longer(col=-Country,
               names_to="Year",
               values_to="Population",
               names_transform=list(Year = as.numeric))
               
dr <- religion %>% select("Country","Religion")
  



  print("inner join of datas:")
join<-inner_join(dc,di,by=c("Country","Year"))%>% 
      inner_join(dp,by=c("Country","Year"))%>%
  inner_join(dr,by="Country")

      join

  
```
```{r,echo=FALSE}

###PART2
##Platting
library(ggplot2)
library(plotly)
library(dplyr)
dat_Iran<- filter(join, Country=="Iran")
print("Dat_Iran:")
dat_Iran


##income plot


ggplot(dat_Iran,aes(x=Year,y=Income))+geom_rect(aes(xmin=1914,xmax=1918,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+geom_rect(aes(xmin=1939,xmax=1945,ymin= -Inf,ymax=Inf),alpha=0.4,fill='grey')+
  geom_point(alpha=0.08,size=4,col="black")+geom_smooth(size=1,col='dark blue')+ theme_bw()+
  labs(x='Year',y='Household income per capita per year[constant$]',title='Household income in Iran')

## fertility plot
ggplot(dat_Iran,aes(x=Year,y=Fertility))+geom_rect(aes(xmin=1914,xmax=1918,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+geom_rect(aes(xmin=1939,xmax=1945,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+
  geom_smooth(size=1.8,col='red')+labs(x="Year",y='Children per woman',title='Fertility in Iran',size=2)


##Iran's neighbors

dat_Iran_region<- filter(join,Country %in%  c('Iran','Afghanistan','Armenia','Iraq','Azerbaijan','Turkey','Turkmenistan','Pakistan'))
print("Dat_Iran_Region:")
dat_Iran_region

#plot neighbors-income-

ggplot(dat_Iran_region,aes(x=Year,y=Income,size=Population))+geom_rect(aes(xmin=1914,xmax=1918,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+geom_rect(aes(xmin=1939,xmax=1945,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+ geom_point(alpha=0.3,aes(col=Country)) +
  labs(x='Year',y='Household income per capita per year[constant$]',title='Household income in Iran and neighbors')
  
#plot neighbors -fertility-
  ggplot(dat_Iran_region,aes(x=Year,y=Fertility,size=Population))+geom_rect(aes(xmin=1914,xmax=1918,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+geom_rect(aes(xmin=1939,xmax=1945,ymin= -Inf,ymax=Inf),alpha=0.7,fill='grey')+ geom_point(alpha=0.3,aes(col=Country))+
  labs(x="Year",y='Children per woman',title='Fertility in Iran and neighbors')

  
library(plotly)
  print("Plotly part")
P <- ggplot(dat_Iran, aes(x=Population, y=Income))+
        geom_point(alpha=0.5,aes(col=Income))+labs(title='Income/Population in Iran',size=0.5 )
ggplotly(P)


allplots <- ggplot(dat_Iran_region, aes(x=Population, y=Income,id=Year,id=Country))+
        geom_point(aes(frame=Year,col=Country,size=Population))+labs(title='Income/Pop in Iran & neighbors')
ggplotly(allplots)

