# A quick and dirty glance

Here, let's just see at a glance what we can do with R very quickly, just to keep focused on the end game in case the first chapters feel a bit long... but hey, you gotta learn how to walk before running!

If you have only one thing to remember, loading data from a 2-columns text file and plotting it is as simple as:
```r
df <- read.table("path/to/data.txt", header=TRUE)
plot(df)
```
_Voilà_, end of the class.

Of course, you can do so much more...

You can define a vector like so:
```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
A <- seq(from=-pi, to=pi, by=.1) # use the sequence function
B <- sin(A) # apply a function to a vector
C <- c(1, 3, 5, 2) # or specify each values
```

And then store these variables into a table (a `data.frame`) with two columns named `x` and `y`:
```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
df1 <- data.frame(x=A, y=B)
head(df1)    # glance at the data
names(df1)   # table's column names
str(df1)     # table's summary
summary(df1) # table's stats
```

Or load data directly from a text file:
```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
# If no header (first line is already data)
df2 <- read.table("Data/rubis_01.txt", header=FALSE)
head(df2) # glance at the data
names(df2) <- c("w", "Intensity") # change column names
# If there is a header (first line is column name)
df3 <- read.table("Data/test.dat", header=TRUE)
head(df3, 2) # glance at the data
```

Plot these:

```{r include=FALSE, warning = FALSE, message=FALSE, cache=FALSE}
par(mar=c(4,4,.5,.5))
```

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE, fig.asp=.8}
# specify x and y
plot(A,B)
# or supply a 2-column data.frame
plot(df2, type="l") # type="l" to plot as lines
# add another plot layer
# as lines
lines(df2$w, df2$Intensity/2, col="red")
# as points
points(df2$w, df2$Intensity/3, col="blue")
# histogram plot
hist(df2$Intensity)
```

And _voilà_! With this, you can already do a lot of tweaking with your data. 

Of course, these plots are ugly and you can do _a lot_ more interesting stuff with R, so let's get to it!



