# Simple variables

In R, the variable attribution is done through the arrow operator `<-` instead of the `=` one – for historical reasons. The equal sign, `=`, would still work though, but you should just take up the habit of using `<-`.

In reality, scalar numbers don't exist in R, they are simply 1 element vector: this is why `[1]` is printed in the following outputs.

To understand what you will encounter further down this class, inputs (what you type in the R script and hit Ctrl+⏎ (⌘+⏎)) are the shaded areas with syntax coloring, while outputs (what R returns) are the shaded are underneath starting with `## [1]`. In case the output spans several lines, each line would start with `## [x]` where `x` is the index of the first element printed on the line.

## Scalars and booleans


Separate in-line instructions with a `;` or just skip a line. 
Different ways of defining a scalar:
```{r, warnings=FALSE}
x <- 1
x
y <- floor(1.2); y
```

You can use the R console as an advanced calculator:
```{r, warnings=FALSE}
1+3
x-y
```

### Booleans
Booleans are handled with the `TRUE` and `FALSE` keywords. Any number non 0 is equivalent to TRUE, 0 is FALSE.
```{r, warnings=FALSE}
z <- TRUE; typeof(z)
as.logical(1)
as.logical(0)
x <- 1; y <- 2
x == y # is equal
x != y # is not equal
x < y  # is smaller than
x <= y # is smaller or equal than
x == y & x < y  # operator "and"
x == y | x < y  # operator "or"
```

### Special values
R handles infinity, NaN, and has $\pi$ defined. Missing numbers are handled through the `NA` keyword (the function `is.na()` checks if the given value contains `NA`).
```{r, warnings=FALSE}
pi
10/0
-10/0
0/0
is.na(NA)
```



### Complex values
R also natively handles complex values:
```{r, warnings=FALSE, error=TRUE}
1+i   # not a valid complex value
1+1i  # valid complex value
exp(1i*pi)
sqrt(-1)
sqrt(-1 + 0i)
Im(exp(1i*pi))
```

## Strings


A string is defined between quotation marks `"string"`. Here are some operations on strings :
```{r, warnings=FALSE, error=TRUE}
phrase <- " Hello World " # definition of a string
paste("phrase =", phrase, ", class(phrase) =", class(phrase))
substr(phrase, 2, 5) # sub-string
paste(tolower(phrase), toupper(phrase), sep=" - ") # change case
sub("o", "a", phrase)  # change the first occurrence of "o" in "a"
gsub("o", "a", phrase) # change all occurrences of "o" in "a"
trimws(phrase)         # trim white spaces
phrase2 <- "1234"
phrase2 - 4321             # won't work: string - double
as.numeric(phrase2) - 4321 # conversion of string to double
```

For more complex operations, see the [stringr](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html) package or the [string cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf).

## Exercises {#exo-variables}


What will be the output of the following entries?



```{r}
x <- "100"; y <- 99.6; 
```


```r
x - ceiling(y)
```

<details>
<summary>Show output</summary>

```{r, warnings=FALSE, error=TRUE, echo=FALSE}
x - ceiling(y)
```

</details>
<br>

```r
1 / sin(as.numeric(x) - ceiling(y))
```

<details>
<summary>Show output</summary>

```{r, warnings=FALSE, error=TRUE, echo=FALSE}
1 / sin(as.numeric(x) - ceiling(y))
```


</details>
<br>

```r
cos(pi/2)==cos(3*pi/2)
```

<details>
<summary>Show output</summary>

```{r, warnings=FALSE, error=TRUE, echo=FALSE}
cos(pi/2)==cos(3*pi/2)
```

</details>
<br>

```r
cos(pi/2)-cos(3*pi/2)
```

<details>
<summary>Show output</summary>

```{r, warnings=FALSE, error=TRUE, echo=FALSE}
cos(pi/2)-cos(3*pi/2)
```

</details>
<br>

```r
all.equal(cos(pi/2),cos(3*pi/2))
```

<details>
<summary>Show output</summary>

```{r, warnings=FALSE, error=TRUE, echo=FALSE}
all.equal(cos(pi/2),cos(3*pi/2))
```

</details>

<br>
<br>
<br>
<br>
<br>

