# Lists

## Definition

Lists allow you to store all types of objects and types of values: booleans, doubles, characters, vectors, other lists, data.frame, etc
```{r, warnings=FALSE}
# initialization
L <- list(name = "John",
          age  = 43,
          kids = list(name=c("Kevin", "Pamela"), # nested list
                      age =c(4,5)
                     )
         )
L
# names of entries (can be changed)
names(L)
# statistics
summary(L)
str(L)
```

## Accessing values and other operations

```{r, warnings=FALSE}
L$name # is a vector
L["age"];typeof(L["age"])     # is a list
L[["age"]];typeof(L[["age"]]) # is a vector
L[[3]]      # is a list (because 'kids' is a list)
L[[3]]['name']   # is a list
L[[3]][['name']] # is a vector
# empty initialization
LL <- list(); LL # no specific size
LL <- vector("list", length=3); LL # specific size
# Concatenation
L1 <- list(wife="Kim", wife.age=38)
L2 <- c(L, L1)
typeof(L2); L2
```

## Exercises {#exo-lists}

- Create a list containing 2 strings, 2 numbers, 2 vectors, 1 list and 2 logical values.
- Give names to the elements in the list.
- Access the first and second elements of the list.
- Add an element at the end of the list.
- Select the second element of a the nested list.
- Remove the second element of the list.
- Create a second list with whatever you want
- Merge the two lists into one list.
- Convert `list(1,2,3,4)` to a vector
- Print the number of objects in the merged list.
- Add a new item `g4 = "Hello"` to the merged list.


