# Vectors

**R is a vectorized language** with built-in arithmetic and relational operators, mathematical functions and types of number. It means that each mathematical function and operator works on a set of data rather than a single scalar value as traditional computer language do, and thus __formal looping can _usually_ be avoided__.

## Different ways of defining a vector

If you want to define a vector by specifying the values, use the function `c()`{.R}, like so:

- Here, `x` is a vector of doubles:
```{r, warnings=FALSE}
x <- c(1, 5, 3, 12, 4.2)
x
```
- But here, `x` is converted to a vector of strings because it contains a string:
```{r, warnings=FALSE}
x <- c(1, 5, 3, "hello")
x
```

- To define a sequence of increasing numbers, either use the notation `start:end` for a sequence going from `start` to `end` by step of 1, or use the `seq()`{.R} function that is more versatile:
```{r, warnings=FALSE}
1:10
seq(-10, 10, by = .5)
seq(-10, 10, length = 6)
seq(-10, 10, along = x)
```
- To repeat values, use the `rep()`{.R} function:
```{r, warnings=FALSE}
rep(0, 10)
rep(c(0, 2), 5)
rep(c(0, 2), each = 5)
```
- To create vectors of random numbers, use `rnorm()`{.R} or `runif()`{.R} for normally or uniformly distributed numbers, respectively:
```{r, warnings=FALSE}
# vector with 10 random values normally distributed around mean 
# with given standard deviation `sd`
rnorm(10, mean=3, sd=1)
# vector with 10 random values uniformly distributed between min and max
runif(10, min = 0, max = 1)
```

## Numerical and categorical data types

Data can be of two different types: numerical, or categorical. Let's say you are measuring a the temperature in a room and recording its value over time: 

```{r}
T1 <- c(22.3, 23.5, 26.0, 30.2)
```

`T1` is a vector containing **numerical** data.

Let's say that now you are recording the temperature level, which can be `low`, `high` or `medium`

```{r}
T2 <- c("low", "low", "medium", "high")
```

`T2` is a vector containing **categorical** data, _i.e._ the data in this example can fall into either of 3 categories. For now, `T2` is however a vector of strings, and we need to tell R that it contains categorical data by using the function `factor()`{.R}:

```{r}
T2 <- factor(T2)
T2
```

We see here that we now have 3 levels, and a numerization of `T2` leads to obtaining the numbers 1, 2 and 3 according to the levels in `T2`:

```{r}
as.numeric(T2)
```

Numerical data can be converted to factors in the same way -- this can be useful sometimes, _e.g._ when plotting with `ggplot` as we will see later:

```{r}
factor(T1)
```

## Principal operations on vectors

### Mathematical operations
Because R is a vectorized language, you don't need to loop on all elements of a vector to perform element-wise operations on it. Let's say that `x <- 1:6`{.R}, then:
```{r, include=FALSE}
x <- 1:6
```

- Addition of a value to all elements:
```{r}
x + 2.5
```
- Multiplication / division of all elements:
```{r}
x*2
```
- Integer division:
```{r}
x %/% 3
```
- Math functions apply on all elements:
```{r}
sqrt(abs(cos(x)))
```
- Power:
```{r}
x^2.5
```
- Multiplication of vectors of the same size is performed element by element:
```{r}
y <- c(2.3, 5, 7, 10, 12, 20)
x*y
```
- Multiplication of vectors of different sizes: the smaller vector is automatically repeated the number of times needed to get a vector of the size of the larger one. It will work also if the longer object length is not a multiple of shorter object length, but the shorter object will be truncated and you'll get an error:
```{r}
x*1:2
x*1:4
```
- Modulo:
```{r}
x %% 2
```
- Outer product of vectors (the result is a matrix):
```{r}
x %o% y
```


### Accessing values
Let's work on this vector `x`:
```{r}
x <- c(5, 3, 4, 9, 3)
```

#### Accessing values by index

- To access values of a vector `x`, use the `x[i]` notation, where `i` is the index you want to access. `i` can be (and in fact, is always) a vector.

:::{.callout-warning}
## Attention
In R, indexes numbering start at **1** !!!
:::

```{r, warnings=FALSE}
# accessing by indexes
x[3]
# access index 1, 5 and 2
x[c(1,5,2)] 
```

- To remove elements `j` from a vector `x`, use the notation `x[-j]`{.R}:
```{r}
# remove elements 1 and 3
x[-c(1,3)]
```

:::{.callout-warning}
## Attention
Writing `x[-c(1,3)]`{.R} will **just print** the result of `x[-c(1,3)]`{.R}, but not actually modify `x`. To really modify `x`, you'd need to write `x <- x[-c(1,3)]`{.R}.
:::

#### Filtering values with tests

You can access values with booleans. Values getting a `TRUE`{.R} will be kept, while values with a `FALSE`{.R} will be discarded:
```{r}
x[c(TRUE,TRUE,TRUE,FALSE,TRUE,TRUE)]
```
Therefore, you can apply tests on your values and filter them very easily:
```{r}
x > 4 # is a vector of booleans
x[x > 4] # is a filtered vector
```

#### Accessing values by name

Finally, values in vectors can be **named**, and thus can be accessed by their name:
```{r, warnings=FALSE}
y <- c(age=32, name="John", pet="Cat")
y
y[c('age','pet')] # prints a named vector
y[['name']] # prints an un-named vector
```

And you can access the names of the vector using `names()`{.R}:
```{r, warnings=FALSE}
names(y)
```


### Sorting, removing duplicates, sampling


- Sorting by ascending number:
```{r, warnings=FALSE}
sort(x)
```
- It works with strings too, but `gtools::mixedsort()`{.R} might be needed for strings mixing letters and numbers:
```{r, warnings=FALSE}
sort(c("c","a","d","ab")) 
sort(c("c", "a10", "a2", "d", "ab"))
gtools::mixedsort(c("c", "a10", "a2", "d", "ab"))
```
- Sorting by descending number:
```{r, warnings=FALSE}
sort(x, decreasing = TRUE) 
```
- Inverting the order of the vector:
```{r}
rev(x)
```

- Find the order of the indexes of the sorting:
```{r, warnings=FALSE}
order(x)
x[order(x)] # is thus equivalent to `sort(x)`
```
- Find duplicates:
```{r, warnings=FALSE}
duplicated(x)
```
- Remove duplicates:
```{r, warnings=FALSE}
unique(x)
```
- Choose 3 random values:
```{r, warnings=FALSE}
sample(x, 3)
```

### Maximum and minimum
This is quite straightforward:
```{r, warnings=FALSE}
# maximum of x and its index
x; max(x); which.max(x) 
# minimum of x and its index
x; min(x); which.min(x)
# range of a vector
range(x)
```

### Characteristics of vectors

- Size of a vector:
```{r, warnings=FALSE}
length(x)
```
- Statistics on vector:
```{r, warnings=FALSE}
summary(x)
```
- Sum of all terms:
```{r, warnings=FALSE}
sum(x)
```
- Average value:
```{r, warnings=FALSE}
mean(x)
```
- Median value:
```{r, warnings=FALSE}
median(x)
```
- Standard deviation:
```{r, warnings=FALSE}
sd(x)
```
- Count occurrence of values:
```{r, warnings=FALSE}
table(x)
```
- Cumulative sum:
```{r}
cumsum(x)
```
- Term-by-term difference
```{r}
diff(x)
```

### Concatenation of vectors
This is done using the `c()` notation: you basically create a vector of vectors, the result being a new vector:
```{r, warnings=FALSE}
# concatenate vectors
z <- c(-1:4, NA, -x); z
```

Another option is to use the `append()`{.R} function that allows for more options:
```{r, warnings=FALSE}
# append values
append(x, 4)    # at the end
append(x, 4:1, 3) # or after a given index
```


## Exercises {#exo-vectors}


Interactive exercises can be found in the [`tutor`](https://colinbousige.github.io/tutor/) package. For this, simply run:

```r
library(tutor)
tuto("vectors")
```

Alternatively, you can just <a href="Data/exo-in-class.zip" download target="_blank">download the archive with all the exercises files</a>, unzip it in your `R class` RStudio project, and edit the R files.

---

<details>
    <summary>**Exercise 1**</summary>

Let's create a named vector containing age of the students in the class, the names of each value being the first name of the students. Then:

- Compute the average age and its standard deviation
- Compute the median age
- What is the maximum, minimum and range of the ages in the class?
- What are all the student names in the class?
- Print the sorted ages by increasing and decreasing order
- Print the ages sorted by alphabetically ordered names (increasing and decreasing)
- Show a histogram of the ages distribution using `hist()`{.R} and play with the parameter `breaks` to modify the histogram
- Show a boxplot of the ages distribution using `boxplot()`{.R}


</details>

<details>
    <summary>**Exercise 2**</summary>

This exercise is adapted from [here](https://perso.esiee.fr/~courivad/R/05-vectors.html#application-population-in-french-cities-1962-2012).

Open Rstudio and create a new R script, save it as `population.R` in your wanted directory, say `Rcourse/`.


<a href="Data/population.csv" download target="_blank">Download</a> the `population.csv` file and save it in your working directory.

A csv file contains raw data stored as plain text and separated by a comma (Comma Separated Values). Open it with Rstudio.

We can of course directly load such file with R and store its data in an appropriate format (_i.e._ a `data.frame`), but this is for the next chapter. For now, just copy-paste the text in the Rstudio script area to:

- Create a `cities` vector containing all the cities listed in `population.csv`
- Create a `pop_1962` and `pop_2012` vectors containing the populations of each city at these years. Print the 2 vectors. 
- Use `names()`{.R} to name values of `pop_1962` and `pop_2012`. Print the 2 vectors again. Are there any change?
- What are the cities with more than 200000 people in 1962? For these, how many residents in 2012?
- What is the population evolution of Montpellier and Nantes?
- Create a `pop_diff` vector to store population change between 1962 and 2012
- Print cities with a negative change
- Print cities which broke the 300000 people barrier between 1962 and 2012
- Compute the total change in population of the 10 largest cities (as of 1962) between 1962 and 2012.
- Compute the population mean for year 1962
- Compute the population mean of Paris over these two years
- Sort the cities by decreasing order of population for 1962

<details>
<summary>Solution</summary>

```{r, warnings=FALSE}
# Create a `cities` vector containing all the cities listed in `population.csv`
cities <- c("Angers", "Bordeaux", "Brest", "Dijon", "Grenoble", "Le Havre", 
            "Le Mans", "Lille", "Lyon", "Marseille", "Montpellier", "Nantes", 
            "Nice", "Paris", "Reims", "Rennes", "Saint-Etienne", "Strasbourg", 
            "Toulon", "Toulouse")
# Create a `pop_1962` and `pop_2012` vectors containing the populations 
# of each city at these years. Print the 2 vectors. 
pop_1962 <- c(115273,278403,136104,135694,156707,187845,132181,239955,
              535746,778071,118864,240048,292958,2790091,134856,151948,
              210311,228971,161797,323724)
pop_2012 <- c(149017,241287,139676,152071,158346,173142,143599,228652,
              496343,852516,268456,291604,343629,2240621,181893,209860,
              171483,274394,164899,453317)
pop_1962; pop_2012
# Use names() to name values of `pop_1962` and `pop_2012`. 
# Print the 2 vectors again. Are there any change?
names(pop_2012) <- names(pop_1962) <- cities
pop_1962; pop_2012
# What are the cities with more than 200000 people in 1962? 
# For these, how many residents in 2012?
cities200k <- cities[pop_1962>200000]
cities200k; pop_2012[cities200k]
# What is the population evolution of Montpellier and Nantes?
pop_2012['Montpellier'] - pop_1962['Montpellier']; pop_2012['Nantes'] - pop_1962['Nantes']
# Create a `pop_diff` vector to store population change between 1962 and 2012
pop_diff <- pop_2012 - pop_1962
# Print cities with a negative change
cities[pop_diff<0]
# Print cities which broke the 300000 people barrier between 1962 and 2012
cities[pop_2012>300000 & pop_1962<300000]
# Compute the total change in population of the 10 largest cities
# (as of 1962) between 1962 and 2012.
ten_largest <- cities[order(pop_1962, decreasing = TRUE)[1:10]]
sum(pop_2012[ten_largest] - pop_1962[ten_largest])
# Compute the population mean for year 1962
mean(pop_1962)
# Compute the population mean of Paris
mean(c(pop_1962['Paris'], pop_2012['Paris']))
# Sort the cities by decreasing order of population for 1962
(pop_1962_sorted <- sort(pop_1962, decreasing = TRUE))
```

</details>
</details>


<br>
<br>
<br>
<br>
<br>


