# Reading/writing all kinds of files


## Reading files

Working in any data-based scientific field, you will encounter many different types of files. ASCII text files are usually predominant, but you may want to read files from coming from Excel, Origin, etc.
Here is a non-exhaustive reminder to help you read the kinds of files you often encounter with R. I invite you to visit the [RopenSci webpage](https://ropensci.org/) for more packages, and if this isn't enough for your need, well, [Google is your friend](http://letmegooglethat.com/?q=r+read+my+file+type).


### Column text files

Most base R functions like `read.csv()`{.R} or `read.table()`{.R} have their `tidyverse` counterpart, like `read_csv()`{.R} and `read_table()`{.R}. I usually prefer the `tidyverse` version as it outputs a `tibble` instead of a `data.frame`. These functions can also be directly provided with an url to the text file.

If your data file is "complicated" -- in the sense that it contains lines or columns to be skipped -- look at the help on the reader function with `?function_name`.

#### Comma separated values

Input file looks like this:
```{bash echo=FALSE}
head Data/tot_population.csv
```

```{r}
library(tidyverse)
read_csv("Data/tot_population.csv")
read_csv("Data/tot_population.csv", skip = 1)
read_csv("Data/tot_population.csv", skip = 1, col_names = LETTERS[1:4])
```

Since version 2.0 of `readr`, `read_csv()`{.R} can also take a vector as argument, which will result in reading all files in the vector successively:

```{r}
read_csv(c("Data/test1.csv",
           "Data/test2.csv"),
         id = "file", show_col_types = FALSE)
```

#### Space separated values

Input file looks like this:
```{bash echo=FALSE}
head Data/rubis_01.txt
```

```r
library(tidyverse)
read_table("Data/rubis_01.txt")
read_table("Data/rubis_01.txt", col_names = c("w","int"))
```


#### Other separators

For tab-separated values, use `read_tsv()`{.R}. For other exotic separators, look into `read_delim()`{.R}.


### Excel files

For this, use the `readxl` library and its function `read_excel()`{.R} returning a `tibble`:

```{r warning=FALSE, message=FALSE}
library(readxl)
read_excel("Data/test.xlsx")
read_excel("Data/test.xlsx", sheet=2) # specify the sheet by its number or its name
```

In case your Excel file contains merged cells, `read_excel()`{.R} will fill the merged cells by `NA` values. If you want to avoid this behavior, use `openxlsx::read.xlsx()`{.R} (which returns a `data.frame`):

```{r warning=FALSE, message=FALSE}
read_excel("Data/test.xlsx", sheet=3)
library(openxlsx)
read.xlsx("Data/test.xlsx", fillMergedCells = TRUE, sheet = 3)
```


### Origin files

If you moved to R coming from a workflow where you used [Origin](https://www.originlab.com/origin), chances are you have some .opj files lying around that you still want to be able to read. Lucky you, the `Ropj` library is here:

```r
library(Ropj)
read.opj("Data/opjfile.opj")
```

### Matlab files

To read Matlab's .mat format datasets, use the [R.matlab package](https://github.com/HenrikBengtsson/R.matlab) and its `readMat()`{.R} function.

```r
library(R.matlab)
df <- readMat("yourfile.mat")
```


### Images

You can read an image as a matrix. For example:

```r
library(png)
readPNG("image.png")
library(tiff)
readTIFF("image.tiff")
```

For more image processing (pictures or videos), I recommend the [`imager` package](https://dahtah.github.io/imager/imager.html).


### Spectroscopic files

In case your spectroscopic data wasn't saved as an ASCII file but as a `spc` or another format, take a look at the [`lightr` package](https://docs.ropensci.org/lightr/), for example.

Example of a reading function for an spc file:

```r
library(tidyverse)
library(lightr)
read_spc <- function(fname){
    d <- lr_parse_spc(fname) # d is a list
    tibble(w         = d[[1]]$wl, 
           intensity = d[[1]]$processed)
}
```


### Compressed binary data files: HDF, netCDF

**HDF**: Go to [this vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/rhdf5/inst/doc/rhdf5.html) to see how to read Hierarchical Data Files.

**netCDF**: see [here](https://www.rdocumentation.org/packages/RNetCDF/versions/2.4-2/topics/read.nc).



## Reading multiple files into a tidy table

We very often encounter the situation where we need to read multiple similar files into a tidy table.
For this, you can use a `for`{.R} loop: this would work but would be un-R-ly, but in some cases you need to do this because you perform some other operations during the `for` loop. 

Let's say we store the list of file names into a vector `file_list` and read these files using the function `read_function()`{.R}:

```r
library(tidyverse)
df <- tibble() # empty initialization
for (file in file_list) {
    df_temp <- read_function(file) %>% 
            mutate(name = file) # add the column `name` to make the tibble tidy
    df <- bind_rows(df, def_temp)
}
```

An R-friendly way of doing this would be to avoid using a `for`{.R} loop:
```r
library(tidyverse)
df <- tibble(name = file_list) %>% 
    mutate(data = map(name, read_function)) %>% 
    unnest(data)
```

In case you are reading csv files, you can use the fact that since the version 2.0 of `readr`, `read_csv()`{.R} takes a vector as first argument. You'll also need to use the `id` argument to get a column with the list of file names:

```r
library(tidyverse)
df <- read_csv(file_list, id = "name")
```


## Writing files

### Text files

Sometimes, you want to output your data as a csv or an Excel file to share it with others or to save your data. Use the `write_csv()`{.R} function to write a csv file (prefer the `tidyverse`'s `write_csv()`{.R} to the base R `write.csv()`{.R} as it is more easy to use):

```r
library(tidyverse)
write_csv(df, "your_file.csv")
```


Note that the `write_*()`{.R} functions will automatically compress outputs if an appropriate extension is given. Three extensions are currently supported: .gz for gzip compression, .bz2 for bzip2 compression and .xz for lzma compression. See the examples in the help for more information.

If you don't want to use csv files, look into `write_tsv()`{.R} for tab-separated values or `write_delim()`{.R} for any delimiter. In case you want to output fixed width files, look into [`gdata`'s `write.fwf()`{.R}](https://www.rdocumentation.org/packages/gdata/versions/2.18.0/topics/write.fwf).


### Excel files

To write Excel files, use the library `openxlsx` and its function `write.xlsx()`{.R} (see the help on the function for more options):

```r
library(openxlsx)
write.xlsx(df, "your_file.xlsx")
```




