# 3D color plots {#colorplots}

```{r include=FALSE, warning=FALSE, message=FALSE}
detachAllPackages <- function() {
  basic.packages <- c("package:stats","package:graphics","package:grDevices","package:utils","package:datasets","package:methods","package:base")
  package.list <- search()[ifelse(unlist(gregexpr("package:",search()))==1,TRUE,FALSE)]
  package.list <- setdiff(package.list,basic.packages)
  if (length(package.list)>0)  for (package in package.list) detach(package, character.only=TRUE)
}
detachAllPackages()
rm(list = ls(all = TRUE))
library(knitr)
```


You may want to plot your data as a color map, like the evolution of a Raman spectrum as a function of temperature, pressure or position. In some cases you'll have a 3-columns `data.frame`{.R} with x, y, and z values (_e.g._ intensity of a peak as a function of the position on the sample), in some cases you can have a list of spectra evolving with a given parameter. 

## The ggplot2 solution

Let's create a dummy set of spectra that we will gather in a tidy `tibble`{.R}.

```{r, message=FALSE}
library(tidyverse)
Nspec <- 40                           # Amount of spectra
N     <- 500                          # Size of the x vector
# Create a fake data tibble
fake_data <- tibble(T = round(seq(273, 500, length=Nspec), 1)) %>% 
    mutate(spec = map(T, ~tibble(w = seq(0, 100, length = N),
                         Intensity = 50*dnorm(w, mean = (./T[1])*20 + 25, 
                                                  sd  = 10+runif(1,max=5)))))
fake_data
fake_data <- fake_data %>% unnest(spec)
fake_data
```

OK, so now we have some fake experimental data stored in a tidy `tibble`{.R} called `fake_data`. We want to plot it as a color map in order to grasp the evolution of the spectra. This can be done through the use of `geom_contour()`{.R} and `geom_contour_filled()`{.R} functions and by providing the `z` aesthetics, or by using the `geom_raster()`{.R} or `geom_tile()`{.R} functions with a `fill` aesthetics. Both methods can be combined, as shown below:

```{r, out.width='100%', warning=FALSE, message = FALSE, fig.asp=.7,cache=FALSE}
# Plotting
colors <- colorRampPalette(c("white","royalblue","seagreen",
                             "orange","red","brown"))
Nbins <- 10
ggplot(data=fake_data, aes(x=w, y=T, z=Intensity)) + 
      geom_contour_filled(bins = Nbins) + 
      ggtitle("Some fake data") + 
      scale_fill_manual(values = colors(Nbins),
                        name = "Intensity\n[arb. units]") +
      labs(x = "Fake Raman Shift [1/cm]",
           y = "Fake Temperature [K]") +
      theme_bw()

ggplot(data=fake_data, aes(x = w, y = T)) + 
      geom_raster(aes(fill = Intensity)) + #geom_tile would work
      geom_contour(aes(z = Intensity), color = "black", bins = 5)+
      ggtitle("Some fake data") + 
      scale_fill_gradientn(colors = colors(10), 
                           name = "Intensity\n[arb. units]") +
      labs(x = "Fake Raman Shift [1/cm]",
           y = "Fake Temperature [K]") +
      theme_bw()
```

Another option is to make a "ridge plot", or a stacking of plots:
```{r, out.width='100%', message = FALSE, fig.asp=.7,cache=FALSE}
colors <- colorRampPalette(c("royalblue","seagreen","orange",
                             "red","brown"))(length(unique(fake_data$T)))
ggplot(data = fake_data, 
       aes(x = w, 
           y = Intensity + as.numeric(factor(T))-1,
           color = factor(T))
       ) + 
    geom_line() + 
    labs(x = "Fake Raman Shift [1/cm]", 
         y = "Fake Intensity [arb. units]") +
    coord_cartesian(xlim = c(25,75)) +
    scale_color_manual(values=colors,name="Fake\nTemperature [K]") +
    theme_bw()
ggplot(data=fake_data, 
       aes(x = w, 
           y = Intensity + as.numeric(factor(T))-1, 
           color = T, 
           group = T)
       )+
    geom_line() + 
    labs(x="Fake Raman Shift [1/cm]", y="Fake Intensity [arb. units]") +
    scale_color_gradientn(colors=colors,name="Fake\nTemperature [K]") +
    coord_cartesian(xlim = c(25,75)) +
    theme_bw()
```



## The base graphics solution

In some cases you end up with a matrix _z_, and two vectors _x_ and _y_. This is easy to plot using the base `image()`{.R} function. For the sake of example, let's just pivot our 3-columns data.frame to such a matrix using `pivot_wider()`{.R}:

```{r, out.width='100%', message = FALSE, warning = FALSE, fig.asp=.6,cache=FALSE}
x <- sort(unique(fake_data$w))
y <- sort(unique(fake_data$T))
z <- as.matrix(fake_data %>% 
                pivot_wider(values_from = Intensity, names_from = T) %>% 
                select(-w)
               )
colors <- colorRampPalette(c("white","royalblue","seagreen","orange","red","brown"))(50)
par(mar = c(4, 4, .5, 4), lwd = 2)
image(x, y, z, col = colors)
```

You can add a legend by using the `image.plot` function:
```{r, out.width='100%', message = FALSE, warning = FALSE, fig.asp=.6,cache=FALSE}
library(fields)
par(mar=c(4, 4, .5, 4), lwd=2)
image.plot(x,y,z, col = colors)
```


## The plotly solution

And finally, if you want to make this an interactive plot, you can use `plot_ly()`{.R}:
```{r, out.width='100%', message = FALSE, warning = FALSE, fig.asp=.6,cache=FALSE}
library(plotly)
aX <- list(title = "Raman Shift [1/cm]")
aY <- list(title = "Temperature [K]")
# Weird but you need to use t(z) here:
z <- t(z)
# Color plot
plot_ly(x = x, y = y, z = z, type = "heatmap", colors = colors) %>% 
   layout(xaxis = aX, yaxis = aY)
```
Or, very cool, an interactive surface plot:
```{r, out.width='100%', message = FALSE, warning = FALSE, fig.asp=.9,cache=FALSE}
plot_ly(x=x, y=y, z=z, type = "surface", colors=colors) %>%
   layout(scene = list(xaxis = aX, yaxis = aY, dragmode="turntable"))
```


## The case of non-regular data

In case you have a set of non-regular data, plotting it as a color map can get tricky: how do we tell the plotting device what color should be in a place where there is no data point?

The solution is to use a spline (or linear, but spline looks usually nicer) interpolation of your 2D data. For this, we can use the `akima` package and its `interp()`{.R} function, like so:


```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
# let's make our data irregular and see the plot is now not working:
fake_data_irreg <- fake_data[sample(nrow(fake_data), nrow(fake_data)/3),]
# let's plot these irregular data
colors <- colorRampPalette(c("white","royalblue","seagreen",
                             "orange","red","brown"))(500)
ggplot(data=fake_data_irreg, aes(x=w, y=T, fill=Intensity)) + 
      geom_raster() + #geom_tile would work
      ggtitle("Some irregular and ugly fake data") + 
      scale_fill_gradientn(colors=colors,name="Intensity\n[arb. units]") +
      labs(x = "Fake Raman Shift [1/cm]",
           y = "Fake Temperature [K]") +
      theme_bw()
# now let's interpolate the data on a regular grid
library(akima)
fake_data_irreg.interp <- with(fake_data_irreg, 
    interp(x = w, y = T, z = Intensity, 
           duplicate="median",
           xo=seq(min(fake_data_irreg$w), max(fake_data_irreg$w), length = 100),
           yo=seq(min(fake_data_irreg$T), max(fake_data_irreg$T), length = 100),
           extrap=FALSE, linear=FALSE)
    )
# fake_data_irreg.interp is a list of 2 vectors and a matrix
str(fake_data_irreg.interp)
# Regrouping this list to a 3-columns data.frame
melt_x <- rep(fake_data_irreg.interp$x, times=length(fake_data_irreg.interp$y))
melt_y <- rep(fake_data_irreg.interp$y, each=length(fake_data_irreg.interp$x))
melt_z <- as.vector(fake_data_irreg.interp$z)
fake_data_irreg.smooth <- na.omit(data.frame(w=melt_x, T=melt_y, Intensity=melt_z))
# Plotting
fake_data_irreg.smooth %>% 
    ggplot(aes(x=w, y=T, fill=Intensity)) + 
        geom_raster() + 
        ggtitle("Some irregular fake data that have been spline-interpolated") + 
        scale_fill_gradientn(colors=colors, name="Intensity\n[arb. units]") +
        labs(x = "Fake Raman Shift [1/cm]", 
             y = "Fake Temperature [K]") +
        theme_bw()
```




## 2D density of points

In case you want to plot a density of points, you have a variety of solutions:
```{r}
df <- tibble(x=rnorm(1e3, mean=c(1,5)),
             y=rnorm(1e3, mean=c(5,1)))
p1 <- ggplot(data=df, aes(x=x,y=y))+ geom_density2d() + ggtitle('geom_density2d()')
p2 <- ggplot(data=df, aes(x=x,y=y))+ geom_hex() + ggtitle('geom_hex()')
p3 <- ggplot(data=df, aes(x=x,y=y))+ geom_bin2d() + ggtitle('geom_bin2d()')
p4 <- ggplot(data=df, aes(x=x,y=y))+ ggtitle('stat_density2d()') +
        stat_density2d(aes(fill = ..density..), geom = "tile", contour = FALSE, n = 200) +
        scale_fill_continuous(low = "white", high = "dodgerblue4")
library(cowplot)
plot_grid(p1,p2,p3,p4)
```


Or the base `smoothScatter()`{.R} function could do the trick:
```{r}
smoothScatter(df)
```





