---
title : "R Exercises - Basic stuff"
format: 
  html:
    toc                   : true
    theme                 : cosmo
    page-layout           : full
    linkcolor             : "#3d54cb"
    code-link             : true
    code-tools            : true
    code-copy             : true
    code-block-border-left: true
    code-block-bg         : true
    self-contained        : true
execute: 
    warning: false
    message: false
    cache: false
---

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
xfun::embed_file("./Archive.zip", text = "Download data files")
```

----

# Exercise 1

- Print the 6 first lines of the R-built-in data.frame `trees`

```{r}
trees
```

- Print only the column names

```{r}

```

- What is the dimension of `trees`?

```{r}

```

- Plot the trees height and volume as a function of their girth in two different graphs. Make sure the axis labels are clear


```{r}

```

- In each graph, add a red dashed line corresponding to the relevant correlation that you observe (average value, linear correlation...)

```{r}

```

- Explain your choice and write the corresponding values (average value and standard deviation, or slope, intercept and corresponding errors). Round the values to 2 decimals.



-------

# Exercise 2 

- Print the 3 first lines of the R-built-in data.frame `USArrests`. This data set contains statistics about violent crime rates by US state. The numbers are given per 100 000 inhabitants, except for `UrbanPop` which is a percentage.

```{r}
USArrests
```

- What is the average murder rate in the whole country?

```{r}

```

- What is the state with the highest assault rate?

```{r}

```

- Create a subset of `USArrests` gathering the data for states with an urban population above (including) 80%.

```{r}

```

- How many states does that correspond to?

```{r}

```

- Within these states, what is the state with the smallest rape rate?

```{r}

```

- Print this subset ordered by decreasing urban population.

```{r}

```

- Print this subset ordered by decreasing urban population and increasing murder rate.

```{r}

```

- Plot an histogram of the percentage of urban population with a binning of 5%. Add a vertical red line marking the average value. Make sure the x axis shows the [0,100] range.

```{r}

```

- Is there a correlation between the percentage of urban population and the various violent crime rates? argument your answer with plots.

```{r}

```
