---
title : "Simules MD Peapods"
author: "C. Bousige"
date  : "`r Sys.Date()`"
output: 
    flexdashboard::flex_dashboard:
        vertical_layout: scroll
        source_code    : embed
bibliography  : "biblio.bib"
csl           : "nature.csl"
link-citations: true
---

```{r setup, include=FALSE, warning = FALSE, cache=FALSE}
library(tidyverse)
library(patchwork)
library(plotly)
```

# Details {data-icon="fa-info-circle"}

<h3>Step 1</h3>


Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 


<h3>Step 2</h3>

Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 


<h3>Step 3</h3>

Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 
Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla Bla bla bla 

# Data acquisition {data-icon="fa-chart-line"}

## Col 1

### The data is acquired here

```{r data_acquisition, message=FALSE, warning=FALSE, echo=TRUE}
raw <- mtcars
```

## Col 2

### The data is treated here [@ashcroft_solid_1976;@fake_article]

```{r data_treat, message=FALSE, warning=FALSE, echo=TRUE}
raw.treated <- raw %>% mutate(mpg_n = mpg/max(mpg))
```


# Plotting the data {data-navmenu="Plotting" data-icon="fa-chart-line"}

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.align="center", fig.height=7}
P <- raw.treated %>% 
    ggplot(aes(mpg_n, disp, color=factor(cyl)))+geom_point()
ggplotly(P, dynamicTicks = TRUE)
```

# Plotting other data {data-navmenu="Plotting" data-icon="fa-chart-line"}

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.align="center", fig.height=7}
z <- volcano
x <- 10 * (1:nrow(z)) # Should that not be ncol?
y <- 10 * (1:ncol(z)) # Should that not be nrow?
plot_ly(x=x, y=y, z=z, type = "surface")
```

# Session Info {data-icon="fa-info-circle"}

```{r}
sessionInfo() %>%
    pander::pander(compact = FALSE)
```


# References

