---
title : "R Exercises - Cognitive Psychology"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
---

<style type="text/css">
blockquote {
  background: #E9F9FF;
  border-left: 5px solid #026086;
  margin: 1.5em 10px;
  padding: 0.5em 10px;
  font-size: 1em;
}
</style>

----

This exercise is based on data from a PhD student in cognitive psychology. In his work, this student collects response times (RT) to two simultaneous tasks, and then he has to analyse the response times.

- At $t_0$, the stimulus S1 is triggered
- At $t_1=t_0 + SOA$, the stimulus S2 is triggered (SOA is the time between the 2 stimuli)
- At $t_2=t_0 + RT1$, the subject responds to stimulus S1
- At $t_3=t_1 + RT2$, the subject responds to stimulus S2

The inter-response interval (IRI) is defined as the time difference between $t_3$ and $t_2$, thus $IRI=SOA+RT2-RT1$. Negatives IRI therefore mean that the response to S2 is emitted before the response to S1.

For theoretical reasons, one can consider that if $SOA=1500$ ms, both stimulis are answered independently. One can thus use the RT1 and RT2 values for SOA=1500 ms to estimate the IRI in the case of independent responses to the stimuli.

---

# Data wrangling

- Download the datafiles <a href="Archive.zip" download target="_blank">archive</a> and unzip it, then create a R project in this folder in Rstudio.
- Load the `tidyverse` and `patchwork` packages:

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- Load the .csv file in the `Data` folder and save it into `raw_data`. Look into the help of `read_csv()`{.R} to help you get rid of the error.

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- Using successive **pipe operations**, we will now create a `dataclean` table from `raw_data`, where we:
    - **Filter** `raw_data` so that `Procedure[Trial]` is only equal to `"EssaisDT"`
    - **Mutate** the table by adding a column `IRI` containing `SOAdur + S2Visuel.RT - S1Audio.RT`
    - **Filter** the table so that some extreme values are excluded:
        - `S1Audio.RT` is smaller or equal to 2500 and larger or equal to 100
        - `S2Visuel.RT` is smaller or equal to 2500 and larger or equal to 100
        - `S1Audio.ACC`, `S1response.ACC` and `S2Visuel.ACC` are equal to 1

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- Let's compute the simulated values in the case the responses to the stimuli are independent, _i.e._ using SOA=1500ms. Using successive pipe operations and starting from `dataclean`:
    - **Filter** rows so that `SOAdur` is equal to 1500
    - Delete the `SOAdur` and `IRI` columns
    - **Mutate** the table to add 3 columns `IRI_sim_xx`, where `xx`=15, 65 or 250 and `IRI_sim_xx = xx + S2Visuel.RT - S1Audio.RT`.
    - Using `pivot_longer()`{.R} and the options `names_to = c(".value", "SOA_dur"), names_pattern = "(.+)_(.+)"`, pivot the columns containing `"IRI_sim_"` into a long table.

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- We want now to get the averaged `IRI` per subject and per SOA, and its standard deviation. Using `group_by()`{.R} and `summarise()`{.R}, store the mean and standard deviation of `IRI` in a table called `stats_obs`, starting from `dataclean`. It should look like this:

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- We want to do the same for the 3 simulated IRI. 
```{r warning = FALSE, message=FALSE, cache=FALSE}

```

# Plotting

- We want now to produce a graph showing the histograms of the observed `IRI` column using `ggplot2`. 
    - Create a `ggplot` using the `dataclean` dataset
    - Set the aesthetics to `x = IRI`
    - Create the histograms using `geom_histogram()`{.R}, with a fill color depending on `SOAdur`
    - Arrange the plots on a grid depending on the `Subject` column using `facet_wrap()`{.R}
    - Add a vertical lign marking the average value for each subject using `geom_vline()`{.R}. The data for these lines are stored in the `stats_obs` dataset.
    - Play with the theme and other ggplot commands to make the plot look like the one below
        - Try plotting a normalized histogram by [looking up on the Internet how to do this](https://www.google.fr/search?source=hp&ei=g0MLXeGwKNLPgweV04-IBA&q=ggplot+normalized+histogram&oq=ggplot+normalized+histogram)

```{r warning = FALSE, message=FALSE, cache=FALSE}

```

- Let's do the same for the simulated dataset. It should look like this:

```{r warning = FALSE, message=FALSE, cache=FALSE}

```