---
title : "R Exercises - Cognitive Psychology"
format: 
  html:
    toc                   : true
    theme                 : cosmo
    page-layout           : full
    linkcolor             : "#3d54cb"
    code-link             : true
    code-tools            : true
    code-copy             : true
    code-block-border-left: true
    code-block-bg         : true
    self-contained        : true
params:
  solution: true
execute: 
    warning: false
    message: false
    cache: false
---

```{r echo=FALSE, warning=FALSE, message=FALSE}
xfun::embed_file("./Archive.zip", text = "Download data files")
```

----

This exercise is based on data from a PhD student in cognitive psychology. In his work, this student collects response times (RT) to two simultaneous tasks, and then he has to analyse the response times.

- At $t_0$, the stimulus S1 is triggered
- At $t_1=t_0 + SOA$, the stimulus S2 is triggered (SOA is the time between the 2 stimuli)
- At $t_2=t_0 + RT1$, the subject responds to stimulus S1
- At $t_3=t_1 + RT2$, the subject responds to stimulus S2

The inter-response interval (IRI) is defined as the time difference between $t_3$ and $t_2$, thus $IRI=SOA+RT2-RT1$. Negatives IRI therefore mean that the response to S2 is emitted before the response to S1.

For theoretical reasons, one can consider that if $SOA=1500$ ms, both stimulis are answered independently. One can thus use the RT1 and RT2 values for SOA=1500 ms to estimate the IRI in the case of independent responses to the stimuli.

---

# Data wrangling

- Download the datafiles <a href="Archive.zip" download target="_blank">archive</a> and unzip it, then create a R project in this folder in Rstudio.
- Load the `tidyverse` and `patchwork` packages:

```{r include=params$solution}
library(tidyverse)
library(patchwork)
```

- Load the .csv file in the `Data` folder and save it into `raw_data`. Look into the help of `read_csv()`{.R} to help you get rid of the error.

```{r include=params$solution}
raw_data <- read_csv2("Data/data exemple.csv", show_col_types = FALSE)
```

- Using successive **pipe operations**, we will now create a `dataclean` table from `raw_data`, where we:
    - **Filter** `raw_data` so that `Procedure[Trial]` is only equal to `"EssaisDT"`
    - **Mutate** the table by adding a column `IRI` containing `SOAdur + S2Visuel.RT - S1Audio.RT`
    - **Filter** the table so that some extreme values are excluded:
        - `S1Audio.RT` is smaller or equal to 2500 and larger or equal to 100
        - `S2Visuel.RT` is smaller or equal to 2500 and larger or equal to 100
        - `S1Audio.ACC`, `S1response.ACC` and `S2Visuel.ACC` are equal to 1

```{r include=params$solution}
dataclean <- raw_data |> 
    filter(`Procedure[Trial]` == "EssaisDT") |>
    mutate(IRI = SOAdur + S2Visuel.RT - S1Audio.RT) |> 
    filter(S1Audio.RT <= 2500 & S1Audio.RT >= 100 &
        S2Visuel.RT <= 2500 & S2Visuel.RT >= 100 &
        S1Audio.ACC == 1 & S1response.ACC == 1 & S2Visuel.ACC == 1)
```

- Let's compute the simulated values in the case the responses to the stimuli are independent, _i.e._ using SOA=1500ms, and save it into a tibble called `IRI_sim`. Using successive pipe operations and starting from `dataclean`:
    - **Filter** rows so that `SOAdur` is equal to 1500
    - Delete the `SOAdur` and `IRI` columns
    - **Mutate** the table to add 3 columns `IRI_sim_xx`, where `xx`=15, 65 or 250 and `IRI_sim_xx = xx + S2Visuel.RT - S1Audio.RT`.
    - Using `pivot_longer()`{.R} and the options `names_prefix = "IRI_sim_", names_to = "SOAdur", values_to = "IRI_sim"`, pivot the columns containing `"IRI_sim_"` into a long table (you need to add the option to select the corresponding columns).

```{r include=params$solution}
IRI_sim <- dataclean |>
    filter(SOAdur == 1500) |>
    select(-c(SOAdur, IRI)) |>
    mutate(
        IRI_sim_15 = 15 + S2Visuel.RT - S1Audio.RT,
        IRI_sim_65 = 65 + S2Visuel.RT - S1Audio.RT,
        IRI_sim_250 = 250 + S2Visuel.RT - S1Audio.RT
    ) |>
    pivot_longer(
        cols = contains("IRI_sim_"),
        names_prefix = "IRI_sim_",
        names_to = "SOAdur",
        values_to = "IRI_sim"
    )
```

- We want now to get the averaged `IRI` per subject and per SOA, and its standard deviation. Using `group_by()`{.R} and `summarise()`{.R}, store the mean and standard deviation of `IRI` in a table called `stats_obs`, starting from `dataclean`. It should look like this:

```{r echo=TRUE}
(stats_obs <- dataclean |>
    group_by(Subject, SOAdur) |>
    summarise(
        mean = mean(IRI),
        sd = sd(IRI)
    ))
```

- We want to do the same for the 3 simulated IRI. 
```{r echo=TRUE}
(stats_sim <- IRI_sim |>
    group_by(Subject, SOAdur) |>
    summarise(
        mean = mean(IRI_sim),
        sd = sd(IRI_sim)
    ))
```

# Plotting

- We want now to produce a graph showing the histograms of the observed `IRI` column using `ggplot2`. 
    - Create a `ggplot` using the `dataclean` dataset
    - Set the aesthetics to `x = IRI`
    - Create the histograms using `geom_histogram()`{.R}, with a fill color depending on `SOAdur`
    - Arrange the plots on a grid depending on the `Subject` column using `facet_wrap()`{.R}
    - Add a vertical lign marking the average value for each subject using `geom_vline()`{.R}. The data for these lines are stored in the `stats_obs` dataset.
    - Play with the theme and other ggplot commands to make the plot look like the one below
        - Try plotting a normalized histogram by [looking up on the Internet how to do this](https://www.google.fr/search?source=hp&ei=g0MLXeGwKNLPgweV04-IBA&q=ggplot+normalized+histogram&oq=ggplot+normalized+histogram)

```{r echo=TRUE}
dataclean |> 
    ggplot(aes(x = IRI, fill = factor(SOAdur))) +
    geom_histogram(aes(y = stat(count / sum(count))), bins=20) +
    scale_y_continuous(labels = scales::percent_format())+
    facet_wrap(~ paste("Subject", Subject)) +
    geom_vline(data = stats_obs, aes(xintercept = mean, color=factor(SOAdur)), lty = 2, show.legend = FALSE) +
    scale_x_continuous(limits = c(-2000,2000))+
    labs(title = "Observations",
         x = "IRI [ms]",
         y = "Occurence", 
         fill="SOA [ms]") +
    theme_bw() +
    theme(strip.background = element_rect(fill = "transparent", colour = NA),
          strip.text = element_text(face = "bold"),
          axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

- Let's do the same for the simulated dataset. It should look like this:

```{r echo=TRUE}
IRI_sim |> 
    ggplot(aes(x = IRI_sim, fill = factor(SOAdur, levels=c(15, 65, 250)))) +
    geom_histogram(aes(y = stat(count / sum(count))), bins = 20) +
    scale_y_continuous(labels = scales::percent_format()) +
    facet_wrap(~ paste("Subject", Subject)) +
    geom_vline(data = stats_sim, aes(xintercept = mean, color=factor(SOAdur)), lty = 2, show.legend = FALSE) +
    scale_x_continuous(limits = c(-2000,2000))+
    labs(
        title = "Simulations",
        x = "IRI [ms]",
        y = "Occurence",
        fill = "SOA [ms]"
    ) +
    theme_bw() +
    theme(
        strip.background = element_rect(fill = "transparent", colour = NA),
        strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    )
```