---
title : "R Exercises - Spectroscopic data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: false
        code_download  : false
params: 
    solution:
        value: false
---

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
library(downloadthis)
download_link(
  link = "https://github.com/colinbousige/rclass/raw/main/Exo/spectro/Archive.zip",
  output_name = "Data Files",
  button_label = "Download Data Files",
  button_type = "default",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```
<br>

----

# Exercise 1

- Load the `readxl` and `tidyverse` libraries.
- Load `FTIR_rocks.xlsx` into a `data.frame` (in fact, `tibble`).
- Rename the columns with simpler names, such as "w", "r1", "r2" and "r3"
- Find the wavenumber value of the maximum of each spectrum
- Create a function `norm01()` that, given a vector, returns the vector normalized to [0,1]
- Normalize all columns of FTIR intensity to [0,1]
- Using base graphics or `ggplot2`, as you wish, try to reproduce the following graphs:

```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE, fig.height=3.6}
library(readxl)
library(tidyverse)
library(ggplot2)
df <- read_excel("Data/FTIR_rocks.xlsx")
names(df) <- c("w", "r1", "r2", "r3")
norm01 <- function(x) {(x-min(x))/(max(x)-min(x))}
df$r1_norm <- norm01(df$r1)
df$r2_norm <- norm01(df$r2)
df$r3_norm <- norm01(df$r3)
ggplot(df, aes(x=w))+
        geom_line(aes(y=r1_norm), size=1, col="black")+
        geom_line(aes(y=r2_norm), size=1, col="royalblue")+
        geom_line(aes(y=r3_norm), size=1, col="red")+
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
ggplot(df, aes(x=w))+
        geom_line(aes(y=r1_norm), size=1, col="black")+
        geom_line(aes(y=r2_norm+1), size=1, col="royalblue")+
        geom_line(aes(y=r3_norm+2), size=1, col="red")+
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
```




# Exercise 2

- Find in the folder "Data" the list of files with the pattern "rubis\_" in their name (check out the `list.files()` function) and store it in `flist`
- What is the length of `flist`?
- Initialize two empty tibbles called `spec` (storing all spectra) and `maxpos` (storing the position of the maximum for each spectrum)
- Using a `for` loop, load each file in `flist` and store them into `spec` by *making it tidy*, and store the position of the maximum of each spectrum in `maxpos`
- Plot the first spectrum in `spec` to see the x range
- Using `ggplot2`, plot the spectra normalized to [0,1] and stacked on top of each other with a vertical shift of 1
- Plot the evolution of the position of the maximum (expected graph below)

```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE, fig.height=6.5}
library(tidyverse)
library(ggplot2)
flist <- list.files(path = "Data", pattern = "rubis")
spec  <- maxpos <- tibble()
for (file in flist){
    df   <- read_table(paste0("Data/",file), col_names=c("x","y"))
    spec <- bind_rows(spec, tibble(x    = df$x,
                                   y    = norm01(df$y),
                                   name = file)
                      )
    m      <- df$x[which.max(df$y)]
    maxpos <- bind_rows(maxpos, tibble(file   = file, 
                                       maxpos = m)
                        )
}
library(patchwork)
p1 <- ggplot(data=spec, aes(x=x,y=y+as.numeric(factor(name))-1,group=name))+
    geom_line()+
    theme_bw()+
    labs(x="Raman Shift [1/cm]", 
         y="Intensity [arb. units]")
p2 <- ggplot(data=maxpos, aes(x=file, y=maxpos)) +
            geom_point(size=2)+
            theme_bw() +
            theme(axis.text.x = element_text(angle = 45, hjust=1))
p1/p2
```

