---
title : "R Exercises - Spectroscopic data - Solution"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: false
        code_download  : true
params: 
    solution:
        value: true
---

# Exercise 1

- Load the `readxl` and `tidyverse` libraries.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
library(readxl)
library(tidyverse)
```

- Load `FTIR_rocks.xlsx` into a `data.frame` (in fact, `tibble`).

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
# Place the Rmd file you are working on in the same folder as the one containing the "Data" folder
# If you do so, the path is relative to where your working Rmd file is located :
df <- read_excel("Data/FTIR_rocks.xlsx")
# Otherwise, you need to provide the whole absolute path
df <- read_excel("/Users/colin/Downloads/Data/FTIR_rocks.xlsx")
```

- Rename the columns with simpler names, such as "w", "r1", "r2" and "r3"

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
names(df) <- c("w","r1","r2","r3")
```

- Find the wavenumber value of the maximum of each spectrum

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
max1 <- df$w[which.max(df$r1)]
max2 <- df$w[which.max(df$r2)]
max3 <- df$w[which.max(df$r3)]
max1;max2;max3
```

- Create a function `norm01()` that, given a vector, returns the vector normalized to [0,1]

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
norm01 <- function(x){
    (x-min(x))/(max(x)-min(x))
}
# check it's working
x <- 1:10
norm01(x)
```

- Normalize all columns of FTIR intensity to [0,1]

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
# First, let's make the tibble tidy
df_tidy <- df %>% pivot_longer(cols=-w,
                    names_to="rock",
                    values_to="intensity")
df_tidy
# now add the new column 'intensity_n' containing the normalized intensities
df_tidy <- df_tidy %>% 
    group_by(rock) %>%
    mutate(intensity_n = norm01(intensity))
df_tidy
```

- Using base graphics or `ggplot2`, as you wish, try to reproduce the following graphs:

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
P1 <- df_tidy %>%
    ggplot(aes(x=w, y=intensity_n, color=rock)) +
        # add the lines with thickness of 1
        geom_line(size=1) +
        # change the axis labels
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]") +
        theme_bw() +
        # set the color to the wanted colors and give a name to 
        # the legend if you want to plot the legend
        # if this line is absent, ggplot uses its default colors (blue, red, green)
        scale_color_manual(values = c("black", "royalblue", "red"), name="Rock:") +
        # to remove the legend, set it to "none"
        # if you remove this line, the default is to show the legend on the right
        theme(legend.position = "top")
P2 <- df_tidy %>%
    ggplot(aes(x=w, 
               y=intensity_n + as.numeric(factor(rock)) - 1, 
               color=rock)) +
        geom_line(size=1)+
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]")+
        theme_bw()+
        theme(legend.position = "none")+
        scale_color_manual(values = c("black", "royalblue", "red"))
library(patchwork)
P1+P2
```



# Exercise 2

- Find in the folder "Data" the list of files with the pattern "rubis\_" in their name (check out the `list.files()` function) and store it in `flist`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
flist <- list.files(path="Data", pattern = "rubis_")
```

- What is the length of `flist`?

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
length(flist)
```

- Initialize two empty tibbles called `spec` (storing all spectra) and `maxpos` (storing the position of the maximum for each spectrum)

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
spec   <- tibble()
maxpos <- tibble(name=flist, max=NA)
```

- Using a `for` loop, load each file in `flist` and store them into `spec` by *making it tidy*, and store the position of the maximum of each spectrum in `maxpos`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
for(file in flist){
    d <- read_table2(file.path("Data",file), col_names = c("w","int"))
    d$name <- file
    maxpos$max[file==flist] <- d$w[which.max(d$int)]
    spec <- bind_rows(spec,d)
}
spec
maxpos
```


Another version of this that does not use a for loop would be like so (try to understand what's happening in this code by running each line one after the other and looking at the help on the functions used):

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
spec <- tibble(name=flist) %>% 
    mutate(data=map(name, ~read_table2(file.path("Data",.), 
                                       col_names = c("w","int")))
          ) %>% 
    unnest(data)
maxpos <- spec %>% 
    group_by(name) %>% 
    summarise(max = w[which.max(int)])
```


- Plot the first spectrum in `spec` to see the x range

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
spec %>% 
    filter(name==flist[1]) %>%
    ggplot(aes(x=w, y=int))+
        geom_line()+
        labs(x="Raman Shift [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
```

- Using `ggplot2`, plot the spectra normalized to [0,1] and stacked on top of each other with a vertical shift of 1

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
spec %>% 
    group_by(name) %>%
    mutate(int_n = norm01(int)) %>%
    ggplot(aes(x=w, 
               y=int_n + as.numeric(factor(name)) - 1,
               group=name))+
        geom_line()+
        labs(x="Raman Shift [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
```

- Plot the evolution of the position of the maximum (expected graph below)

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
maxpos %>%
    ggplot(aes(x=name, y=max))+
        geom_point()+
        theme_bw()+
        theme(axis.text.x = element_text(angle = 45, hjust=1))
```

