---
title : "R Exercises - TGA data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
params: 
    solution:
        value: FALSE
---


- Load the `tidyverse` package
- Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
- Load it into a `tibble`.
- Plot the mass (column `Value`) as a function of the temperature `Tr` (Temperature Read) with lines, using `ggplot2`
    + Remove data from the temperature decrease
    + Define nice axis labels with the units
    + Convert degrees Celsius to Kelvins
- Write a function returning the derivative $\partial y/\partial x$ given two vectors x and y
- Add the temperature derivative of the mass reading in a red dashed line in a panel below the previous graph
    + Zoom on the data to see the variations
    + Remove the x label of the top panel and have the same x axis for both panels
    + Try to reproduce the following graph:

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
library(tidyverse)
# Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
# Load it into a `tibble`.
d <- read_table2("Data/ATG.txt", 
                skip=12,
                col_names=c("Index", "t", "Ts", "Tr", "Value"), 
                n_max=4088)
# Plot the mass as a function of the temperature `Ts` with lines, using base graphics and then `ggplot2`
# Remove data from the temperature decrease
d <- d %>% filter(t < t[which.max(Tr)])
# Define nice axis labels with the units
# Convert degrees Celsius to Kelvins
P <- d %>% 
    ggplot(aes(x=Tr+273.15, y=Value)) +
        geom_line() +
        labs(x="T [K]", y="Mass [mg]") +
        theme_bw()
P
# Write a function returning the derivative $\partial y/\partial x$ given two vectors x and y
derivative <- function(x, y, n=1) {
    # returns d^ny/dx^n
    d <- tibble(x=x, y=y)
    for (i in 1:n) {
        x  <- d$x
        y  <- d$y
        dy <- diff(y)/diff(x)# the actual derivative
        dx <- x[-length(x)]+diff(x)/2# centers the X values
        d  <- tibble(x=dx,y=dy)
    }
    d
}
dd   <- derivative(d$Tr,d$Value)
dd$x <- dd$x + 273 # convert to Kelvins
dd
# find the maximun of the derivative to find the inflexion point
# the first values are infinite, we want the maximum around 630
xmax <- dd %>% filter(x>400) 
xmax <- xmax$x[which.min(xmax$y)]
xmax
# Add the derivative in a red dashed line in a panel below the previous graph
# Zoom on the data to see the variations
# Remove the x label of the top panel and have the same x axis for both panels
p1 <- P + geom_vline(xintercept = xmax, lty=2, alpha=.5)+
         theme_bw() +
         theme(axis.title.x=element_blank(),
               axis.text.x=element_blank())
p2 <- ggplot(data=dd, aes(x=x, y=y), color="red") +
         geom_line(col="red", lty=2) +
         ylim(c(-.15,.05))+
         labs(x="T [K]", y="Derivative") +
         geom_vline(xintercept = xmax, lty=2, alpha=.5)+
         theme_bw()
```


```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE}
library(patchwork)
p1/p2+ plot_layout(heights = c(4, 1))
```