---
title : "R Exercises - Spectroscopic data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: false
        code_download  : TRUE
---



# Exercise 1 - FTIR

- Load the `readxl` and `tidyverse` libraries.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Load `FTIR_rocks.xlsx` into a `data.frame` (in fact, `tibble`).

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Rename the columns with simpler names, such as "w", "r1", "r2" and "r3"

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Find the wavenumber value of the maximum of each spectrum and store them

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Create a function `norm01()` that, given a vector, returns the vector normalized to [0,1]

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Normalize all columns of FTIR intensity to [0,1]

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Using base graphics or `ggplot2`, as you wish, try to reproduce the following graphs:


```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```






# Exercise 2 - Raman

- Find in the folder "Data" the list of files with the pattern "rubis\_" in their name (check out the `list.files()` function) and store it in `flist`
 
```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- What is the length of `flist`?

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Initialize an empty list called `spec` and an empty `data.frame` called `maxpos`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Using a `for` loop, load each file in `flist` into separate elements of `spec`, and store the position of the maximum of each spectrum in `maxpos`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Plot the first element of `spec` to see the x range

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Using a `for` loop and base graphics, plot the spectra normalized to [0,1] and stacked on top of each other with a vertical shift of 1
    - Try doing the same with `ggplot2` (expected graph below)

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Plot the evolution of the position of the maximum (expected graph below)


```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

