---
title : "R Exercises - TGA data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
params: 
    solution:
        value: FALSE
---



- Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
- Load it into a `data.frame`.
- Plot the mass as a function of the temperature `Ts` with lines, using base graphics and then `ggplot2`
    + Remove data from the temperature decrease
    + Define nice axis labels with the units
    + Add `mass` vs. `Tr` as a blue dashed line
    + Convert degrees Celsius to Kelvins
- Write a function returning the derivative of a `data.frame` 
    + Add the derivative in a red dashed line in a panel below the previous graph
    + Zoom on the data to see the variations
    + Remove the x label of the top panel and have the same x axis for both panels
    + Try to reproduce the following graph:

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
# Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
# Load it into a `data.frame`.
d <- read.table("Data/ATG.txt", 
                skip=12,
                header=FALSE, 
                nrows=4088)
names(d) <- c("Index", "t", "Ts", "Tr", "Value")
head(d)
# Plot the mass as a function of the temperature `Ts` with lines, using base graphics and then `ggplot2`
# Remove data from the temperature decrease
d <- subset(d, t < d$t[which.max(d$Ts)])
# Define nice axis labels with the units
# Add `mass` vs. `Tr` as a blue dashed line
# Convert degrees Celsius to Kelvins
plot(d$Ts+273.15, d$Value, 
    type="l", 
    xlab="T [K]", 
    ylab="Mass [mg]")
lines(d$Tr+273.15, d$Value, col="blue", lty=2)
library(ggplot2)
ggplot(data=d, aes(x=Ts+273.15, y=Value)) +
     geom_line() +
     geom_line(aes(x=Tr+273.15, y=Value), col="blue", lty=2) +
     labs(x="T [K]", y="Mass [mg]") +
     theme_bw()
# Write a function returning the derivative of a `data.frame` 
derivative <- function(x, y, n=1) {
  # returns d^ny/dx^n
  d <- data.frame(x=x,y=y)
  for (i in 1:n) {
    x  <- d$x
    y  <- d$y
    dy <- diff(y)/diff(x)# the actual derivative
    dx <- x[-length(x)]+diff(x)/2# centers the X values
    d  <- data.frame(x=dx,y=dy)
  }
  d
}
dd <- derivative(d$Ts,d$Value)
# Add the derivative in a red dashed line in a panel below the previous graph
# Zoom on the data to see the variations
# Remove the x label of the top panel and have the same x axis for both panels
# ---- base graphics
M  <- matrix(c(1,2), byrow=TRUE, nrow=2)
nf <- layout(M, heights=c(1,.5), widths=c(1))
par(cex.lab=1.5, cex.axis=1.5, mgp = c(3, 0.5, 0), 
    tck=0.02, mar=c(.5, 4.5, 1, 1), las=1, lwd=2)
plot(d$Ts+273.15, d$Value, 
    xlim=c(300, 1200),
    type="l", 
    axes=FALSE,
    xlab="T [K]", 
    ylab="Mass [mg]")
axis(2);axis(4, labels=FALSE);
axis(1, labels=FALSE);axis(3, labels=FALSE);
box()
par(mar=c(4.5, 4.5, 0, 1))
plot(dd$x+273.15, dd$y, 
    ylim=c(-.15,.05),
    xlim=c(300, 1200),
    col="red", 
    type="l",
    axes=FALSE,
    lty=2, 
    xlab="T [K]", ylab="Derivative")
axis(1, labels=TRUE, tck=0.06)
axis(2, labels=TRUE, tck=0.06)
axis(3, labels=FALSE, tck=0.06)
axis(4, labels=FALSE, tck=0.06)
box()
# ---- ggplot2
xmax <- dd[dd$x>400-273,]
xmax <- xmax$x[which.max(abs(xmax$y))]+273+3
p1 <- ggplot(data=d, aes(x=Ts+273, y=Value)) +
         geom_line() +
         labs(x="", y="Mass [mg]") +
         ylim(c(24,31))+
         geom_vline(xintercept = xmax, lty=2, alpha=.5)+
         scale_x_continuous(limits=c(310,1100),
                            breaks = seq(0,1100,100), 
                            labels=seq(0,1100,100))+
         theme_bw() +
         theme(axis.text.x=element_blank(),
               plot.margin=unit(c(0.5,0.5,0,0.5), "pt"))
p2 <- ggplot(data=dd, aes(x=x+273, y=y), color="red") +
         geom_line(col="red", lty=2) +
         ylim(c(-.15,.05))+
         labs(x="T [K]", y="Derivative") +
         geom_vline(xintercept = xmax, lty=2, alpha=.5)+
         scale_x_continuous(limits=c(310,1100),
                            breaks = seq(0,1100,100), 
                            labels=seq(0,1100,100))+
         theme_bw()+
         theme(plot.margin=unit(c(-10,0.5,0.5,0.5), "pt"))
```


```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE}
library(patchwork)
p1/p2+ plot_layout(heights = c(4, 1))
```