---
title : "R Exercises - Basic stuff"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
params: 
    solution:
        value: FALSE
---


```{r setup, include=FALSE}
library(knitr)
knit_engines$set(asis = function(options) {
  if (options$echo && options$eval) knit_child(text = options$code)
})
```

Set up the environment:

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}
library(ggplot2)
library(tidyverse)
theme_set(theme_bw())
```

-------

# Exercise 1

- Print the 6 first lines of the R-built-in data.frame `trees`

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
head(trees,6)
```

- Print only the column names

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
names(trees)
```

- What is the dimension of `trees`?

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
dim(trees)
```

- Plot the trees height and volume as a function of their girth in two different graphs. Make sure the axis labels are clear

- In each graph, add a red dashed line corresponding to the relevant correlation that you observe (average value, linear correlation...)

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
# base R solution
plot(trees$Girth, trees$Height, xlab='Girth', ylab='Height')
abline(col='red',lty=2,h=mean(trees$Height))
plot(trees$Girth, trees$Volume, xlab='Girth', ylab='Volume')
fit <- lm(trees$Volume~trees$Girth)
abline(col='red',lty=2, coef(fit))
# ggplot solution
ggplot(data=trees, aes(Girth, Height))+
    geom_point()+geom_hline(col='red',yintercept=mean(trees$Height),lty=2)
ggplot(data=trees, aes(Girth, Volume))+
    geom_point()+geom_smooth(col='red',method="lm",lty=2)
```

- Explain your choice and write the corresponding values (average value and standard deviation, or slope, intercept and corresponding errors). Round the values to 2 decimals.

```{asis include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
The average height is `r mean(trees$Height)` with standard deviation `r round(sd(trees$Height),2)`.

The volume evolves with a slope `r round(coef(fit)[2],2)` ± `r round(summary(fit)$coef[2,'Std. Error'],2)` and intercept `r round(coef(fit)[1],2)` ± `r round(summary(fit)$coef[1,'Std. Error'],2)`.
```

-------

# Exercise 2 

- Print the 3 first lines of the R-built-in data.frame `USArrests`. This data set contains statistics about violent crime rates by US state. The numbers are given per 100 000 inhabitants, except for `UrbanPop` which is a percentage.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
head(USArrests,3)
```

- What is the average murder rate in the whole country?

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
mean(USArrests$Murder)
```

- What is the state with the highest assault rate?

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
row.names(USArrests)[which.max(USArrests$Assault)]
```

- Create a subset of `USArrests` gathering the data for states with an urban population above (including) 80%.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
USArrests80 <- subset(USArrests, UrbanPop>=80)
# or
USArrests80 <- USArrests %>% filter(UrbanPop>=80)
```

- How many states does that correspond to?

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
nrow(USArrests80)
```

- Within these states, what is the state with the smallest rape rate?

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
row.names(USArrests80)[which.min(USArrests80$Rape)]
```

- Print this subset ordered by decreasing urban population.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
USArrests80[order(-USArrests80$UrbanPop),]
```

- Print this subset ordered by decreasing urban population and increasing murder rate.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
USArrests80[order(-USArrests80$UrbanPop, USArrests80$Murder),]
```

- Plot an histogram of the percentage of urban population with a binning of 5%. Add a vertical red line marking the average value. Make sure the x axis shows the [0,100] range.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
hist(USArrests$UrbanPop, breaks=seq(0,100,5))
abline(v=mean(USArrests$UrbanPop), col='red', lwd=3)
# ggplot solution
ggplot(data=USArrests, aes(x=UrbanPop))+
    geom_histogram(breaks=seq(0,100,5),color="black", alpha=.2)+
    geom_vline(xintercept = mean(USArrests$UrbanPop), col='red')
```

- Is there a correlation between the percentage of urban population and the various violent crime rates? argument your answer with plots.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
par(mfrow=c(2,2),mar=c(4,4,1,1))
plot(USArrests$UrbanPop,USArrests$Murder, pch=16)
plot(USArrests$UrbanPop,USArrests$Assault, pch=16)
plot(USArrests$UrbanPop,USArrests$Rape, pch=16)
```

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
p1 <- ggplot(data=USArrests, aes(x=UrbanPop, y=Murder))+geom_point()
p2 <- ggplot(data=USArrests, aes(x=UrbanPop, y=Assault))+geom_point()
p3 <- ggplot(data=USArrests, aes(x=UrbanPop, y=Rape))+geom_point()
library(patchwork)
p1+p2+p3
```

```{asis include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
No clear correlation appears.
```