---
title : "R Exercises - Spectroscopic data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: false
        code_download  : false
params: 
    solution:
        value: false
---

# Exercise 1

- Load the `readxl` and `tidyverse` libraries.
- Load `FTIR_rocks.xlsx` into a `data.frame` (in fact, `tibble`).
- Rename the columns with simpler names, such as "w", "r1", "r2" and "r3"
- Find the wavenumber value of the maximum of each spectrum
- Create a function `norm01()` that, given a vector, returns the vector normalized to [0,1]
- Normalize all columns of FTIR intensity to [0,1]
- Using base graphics or `ggplot2`, as you wish, try to reproduce the following graphs:

```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE, fig.height=3.6}
library(readxl)
library(tidyverse)
df <- read_excel("Data/FTIR_rocks.xlsx")
names(df) <- c("w", "r1", "r2", "r3")
norm01 <- function(x) {(x-min(x))/(max(x)-min(x))}
df$r1_norm <- norm01(df$r1)
df$r2_norm <- norm01(df$r2)
df$r3_norm <- norm01(df$r3)
ggplot(df, aes(x=w))+
        geom_line(aes(y=r1_norm), size=1, col="black")+
        geom_line(aes(y=r2_norm), size=1, col="royalblue")+
        geom_line(aes(y=r3_norm), size=1, col="red")+
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
ggplot(df, aes(x=w))+
        geom_line(aes(y=r1_norm), size=1, col="black")+
        geom_line(aes(y=r2_norm+1), size=1, col="royalblue")+
        geom_line(aes(y=r3_norm+2), size=1, col="red")+
        labs(x="Wavenumber [1/cm]", y="Intensity [arb. units]")+
        theme_bw()
```




# Exercise 2

- Find in the folder "Data" the list of files with the pattern "rubis\_" in their name (check out the `list.files()` function) and store it in `flist`
- What is the length of `flist`?
- Initialize an empty list called `spec` and an empty `data.frame` called `maxpos`
- Using a `for` loop, load each file in `flist` into separate elements of `spec`, and store the position of the maximum of each spectrum in `maxpos`
- Plot the first element of `spec` to see the x range
- Using a `for` loop and base graphics, plot the spectra normalized to [0,1] and stacked on top of each other with a vertical shift of 1
    - Try doing the same with `ggplot2` (expected graph below)
- Plot the evolution of the position of the maximum (expected graph below)

```{r echo=FALSE, warning = FALSE, message=FALSE, cache=FALSE, fig.height=6.5}
flist <- list.files(path = "Data", pattern = "rubis")
spec <- list()
specdf <- tibble()
maxpos <- data.frame()
for (file in flist){
    df <- as_tibble(read.table(paste0("Data/",file), col.names=c("x","y")))
    spec[[file]] <- df
    specdf <- rbind(specdf, tibble(x=df$x,y=norm01(df$y),name=file))
    x <- spec[[file]][[1]]
    y <- spec[[file]][[2]]
    m <- x[which.max(y)]
    maxpos <- rbind(maxpos, data.frame(file=file, maxpos=m))
}
# plot(NA,NA,
#     xlim=c(3100,3500),
#     ylim=c(0,length(flist)),
#     xlab="Raman Shift [1/cm]",
#     ylab="Intensity [arb. units]"
#     )
# i <- 0
# for (file in flist){
#     x <- spec[[file]][,1]
#     y <- spec[[file]][,2]
#     lines(x, norm01(y)+i)
#     i <- i+1
# }
library(cowplot)
p1 <- ggplot(data=specdf, aes(x=x,y=y+as.numeric(factor(name))-1,group=name))+
    geom_line()+theme_bw()+labs(x="Raman Shift [1/cm]", y="Intensity [arb. units]")
p2 <- ggplot(data=maxpos, aes(x=file, y=maxpos)) +
            geom_point(size=2)+
            theme_bw() +
            theme(axis.text.x = element_text(angle = 45, hjust=1))
plot_grid(p1,p2,align = 'v', axis = 'l', nrow=2)
```