---
title : "R Exercises - TGA data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
---


- Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
- Load it into a `data.frame`.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```


- Plot the mass as a function of the temperature `Ts` with lines, using base graphics and then `ggplot2`
    + Remove data from the temperature decrease
    + Define nice axis labels with the units
    + Add `mass` vs. `Tr` as a blue dashed line
    + Convert degrees Celsius to Kelvins

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```


- Write a function returning the derivative of a `data.frame` 
    + Add the derivative in a red dashed line in a panel below the previous graph
    + Zoom on the data to see the variations
    + Remove the x label of the top panel and have the same x axis for both panels
    + Try to reproduce the following graph:


```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

