---
title : "R Exercises - G(r,t)"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc            : true
        toc_float      : true
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
---

# A bit of context

In this exercise, we will see how to make a color plot of $4\pi r^2\times G(r,t)$ data obtained from Molecular Dynamics simulations.
These files can be downloaded <a href="Data/Grt.zip" download target="_blank">here</a>.
These auto-correlators $G(r,t)$ are obtained from MD trajectories of a fluid encapsulated in a porous medium by computing the probability that each molecule moves by a distance $r$ over a time $t$. The MD trajectories are performed with [LAMMPS](https://lammps.sandia.gov/) and the trajectories treated by a home-made C program producing these files.

In these simulations, we varied the depth of the interaction well. It is given in the name of the file such as `Grt_xx.dat`, where xx = 0.01, 0.1, 0.2, 0.3, 0.5, 0.8, 1 corresponds to $\varepsilon/k_BT$. $\varepsilon/k_BT$ is the depth of the interaction well in units of temperature.

The `Grt_xx.dat` files contain the $4\pi r^2\times G(r,t)$ data under the form of a matrix, with $r$ increasing with the rows and $t$ with the columns. $r$ goes from 0 to 20 Å and $t$ goes from 0 to 200 ps.

# Reading and tidying the data

We want to read all data files and store them into a tidy tibble for ease of plotting.

- Load the `tidyverse` and `ggplot2` packages

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Find all files starting by "Grt" and store them in a vector `Grt_files`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Initiate an empty tibble `Grt` that will store all data

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Using a `for` loop:
    - read the files
    - add columns names as the times values
    - add the `r.A` column containing the r values
    - and recursively (using the pipe `%>%`) 
        - pivot the table to make it tidy, with the columns `r.A`, `t.ps` and `r2Grt`
        - make sure that the times are read as numeric values
        - add the `epskt` column containing the $\varepsilon/k_BT$ value
    - store this in `Grt`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```




# Plotting data


We will now plot this as a colored heatmap using `geom_raster` from `ggplot2`:

- We will first plot `Grt` only for times below 100 ps and $\varepsilon/k_BT=0.5$, so recursively filter the data accordingly, then provide this to `ggplot`

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```


+ Use a nicer axis labels

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

+ Do you see anything on the plot?
+ In fact, we need to saturate values above a certain limit in order to see the fine evolution of the $G(r,t)$ auto-correlator. Add a mutation of the tibble attributing the limit value to all points above this limit, and play with the limit value in order to have a nicer plot.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Add a vertical dashed line in x=8.18Å marking the mean pore size for this structure.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Use a nicer color scheme with more colors in order to better see the variations.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

- Now we want to see the evolution of $G(r,t)$ as a function of $\varepsilon/k_BT$. Using the procedure above, plot $G(r,t)$ for all 
$\varepsilon/k_BT$ on a grid.

```{r include=TRUE, warning = FALSE, message=FALSE, cache=FALSE}

```

