---
title : "R Exercises - TGA data"
date  : "`r Sys.Date()`"
output: 
    html_document:
        toc_depth      : 4
        highlight      : tango
        number_sections: true
        code_download  : FALSE
params: 
    solution:
        value: FALSE
---

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align="center"}
library(downloadthis)
download_link(
  link = "https://github.com/colinbousige/rclass/raw/main/Exo/tga/Archive.zip",
  output_name = "Data Files",
  button_label = "Download Data Files",
  button_type = "default",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```
<br>

----

- Load the `tidyverse` package
- Download the TGA data file <a href="Data/ATG.txt" download target="_blank">ATG.txt</a>
- Load it into a `tibble`.

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
library(tidyverse)
d <- read_table("Data/ATG.txt", 
                skip=12,
                col_names=c("Index", "t", "Ts", "Tr", "Value"), 
                n_max=4088)
```

- Plot the mass (column `Value`) as a function of the temperature `Tr` (Temperature Read) with lines, using `ggplot2`
    + Remove data from the temperature decrease
    + Define nice axis labels with the units
    + Convert degrees Celsius to Kelvins

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
d_filtered <- d %>% 
        filter(t < t[which.max(Tr)], # Remove data from the temperature decrease
               Tr > min(Tr)) %>%     # Remove repeated data at the beginning
        mutate(T_K = Tr + 273.15)    # Convert degrees Celsius to Kelvins
# Define nice axis labels with the units
p0 <- d_filtered %>% 
    ggplot(aes(x = T_K, y = Value)) +
        geom_line() +
        labs(x="T [K]", y="Mass [mg]") +
        theme_bw()
p0
```

- Write a function returning the derivative $\partial y/\partial x$ given two vectors x and y

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
# first derivative
derivative <- function(x, y) {
    # returns dy/dx
    dy <- diff(y)/diff(x)           # the actual derivative
    dx <- x[-length(x)] + diff(x)/2 # centers the X values
    tibble(x = dx, y = dy)
}
# alternatively, derivative at any order:
derivative <- function(x, y, n=1) {
    # returns d^ny/dx^n
    d <- tibble(x = x, y = y)
    if(n>0){
        for (i in seq_along(n)) {
            dy <- diff(d$y)/diff(d$x)             # the actual derivative
            dx <- d$x[-length(d$x)] + diff(d$x)/2 # centers the X values
            d  <- tibble(x = dx, y = dy)
        }
    }
    d
}
```

- Add the temperature derivative of the mass reading in a red dashed line in a panel below the previous graph
    + Zoom on the data to see the variations
    + Remove the x label of the top panel and have the same x axis for both panels
    + Try to reproduce the following graph:

```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
# find the minimun of the derivative to find the inflexion point
dd   <- derivative(d_filtered$T_K, d_filtered$Value)
xmin <- dd$x[which.min(dd$y)]
xmin
# Add the derivative in a red dashed line in a panel below the previous graph
# Zoom on the data to see the variations
p2 <- ggplot(data=dd, aes(x=x, y=y), color="red") +
         geom_line(col="red", lty=2) +
         ylim(c(-.15,.05))+
         labs(x="T [K]", y="Derivative") +
         geom_vline(xintercept = xmin, lty=2, alpha=.5)+
         theme_bw()
# Remove the x label of the top panel and have the same x axis for both panels
p1 <- p0 + geom_vline(xintercept = xmin, lty=2, alpha=.5)+
         theme_bw() +
         theme(axis.title.x=element_blank(),
               axis.text.x=element_blank())
```


```{r include=params$solution, warning = FALSE, message=FALSE, cache=FALSE}
library(patchwork)
p1/p2+ plot_layout(heights = c(4, 1))
```