---
title : "Typical errors"
date  : "`r Sys.Date()`"
output: 
    bookdown::pdf_document2: 
        toc            : false
        highlight      : tango
        theme          : cosmo
---

# Installing/Loading libraries

**Do not** leave a `install.packages()` in a Rmd file: otherwise you will install it every time you knit the document. Do it once, then just load the library once.

```{r warning=FALSE, error=TRUE}
install.packages(tidyverse) # put the quotes in install.packages()
library(tidyverse) #no quotes required in library()
```


# Reading files

**Always** prefer the `tidyverse` version that directly returns a tibble and better reads the column names:


| file    | Base R    | tidyverse  |
|:---:|:---:|:---:|
| .csv    |  `read.csv()`{.R}   |  `read_csv()`{.R}   |
| .tsv    |  `read.tsv()`{.R}   |  `read_tsv()`{.R}   |
| .txt/.dat (space separation)    |  `read.table()`{.R}   |  `read_table2()`{.R}   |


```r
confirmed_raw <- tibble(read.csv(url("https://....")))
colnames(confirmed_raw) <- gsub("X","",names(confirmed_raw), fixed = TRUE)
# REPLACE WITH
confirmed_raw <- read_csv("https://....")
```

# Changing a variable


Prints the results but **no change** to the variable `df`:
```r
df %>% filter(x>0)
```

**Changes** `df`:
```r
df <- df %>% filter(x>0)
```

# Functions


Functions return **the last thing that is called**.

```r
myfunc <- function(x){
    y <- x^2
}#returns nothing
myfunc <- function(x){
    y <- x^2
    y
}#returns y
myfunc <- function(x){
    y <- x^2
    return(y)
}#returns y
myfunc <- function(x){
    y <- x^2
    if(x>1){
        y+2
    }
    z <- x+y
    z
}#returns z
myfunc <- function(x){
    y <- x^2
    if(x>1){
        return(y+2) #stops here
    }
    z <- x+y
    z
}#returns y+2 or z depending on x
```

So to make a function returning operations on a data.frame, you can do this (but see how **you still have to save it**):

```{r}
myfunc <- function(df, min){
    df %>% filter(x > min)
}
df <- tibble(x=-2:2)
df2 <- myfunc(df, 0)
df
df2
```




# Hardcoding

**Do not** do stuff like that :

```r
# Remove the lines from the Diamond Princess
df <- df[-89,]
# Data for cases above the minimum in China
alldata <- alldata %>% filter(confirmed >=521 )
# Where does this 521 value come from ???
```

But **always** prefer explicit operations:

```r
#Remove the lines from the Diamond Princess
df <- df %>% filter(country!="Diamond Princess")
# Data for cases above the minimum in China
minChina <- alldata %>% 
                filter(country == 'China') %>% 
                select(confirmed) %>% 
                min()
alldata <- alldata %>% filter(confirmed >= minChina)
```
